Contribution Title\thanks{Supported by organization x.}

### 2019-10-24 paper/paper.tex
Updated intro and motivation

% Motivate why unambiguous parsing is a good (and important) idea.
Ever since the early 60s, it has been known that determining whether a context-free
grammar is ambiguous is undecidable~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
a large number of restricted grammars have been developed to guarantee
that language definitions are unambiguous. This traditional view of
only allowing unambiguous grammars has been taken for granted as the
only truth: \emph{the} way of how the syntax of a language must be
defined ~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
%
However, recent work by
Palmkvist and Broman~\cite{palmkvistCreatingDomainSpecificLanguages2019} shows why
carefully designed ambiguous syntax definitions are in some cases
preferable compared to completely unambiguous
definitions. Specifically, instead of forcing language designers to
make more or less arbitrary choices to disambiguate the language, a more natural
solution may be to leave the decision up to the programmer.

@@

This paper focuses on the problems that arise if a grammar definition
is not guaranteed to be unambiguous. Concretely, if parsing a program
could result in several correct parse trees, the grammar is obviously
ambiguous. The programmer then needs to disambiguate the program by,
for instance, inserting extra parentheses. We say that a program is
\emph{resolvably ambiguous} if the ambiguity can be resolved by
modifying the program, such that each of the possible parse trees can
be selected using different modifications.
%
We can divide the problem of resolvable ambiguity into two
different decision problems.
%
The \emph{dynamic resolvable ambiguity problem} is as follows:
given a parse function for a language, and a specific program that
is parsed into a set of parse trees, decide whether it is possible to
rewrite the program, such that each parse tree can be generated
unambiguously.
%
This problem can be generalized into the \emph{static resolvable
  ambiguity problem}: given a parse function, decide whether all
parsable programs are resolvably ambiguous.
%
Static resolvability means that a language designer can be
confident that the end-user (the programmer) can always resolve an
ambiguity by modifying the program. Identifying such ambiguities
for a \emph{specific} program and suggesting disambiguations is
enabled by dynamic resolvability.
%
Static resolvability gives formally verified guarantees on a
syntax definition, whereas dynamic resolvability enables the
language designer to test the resolvability of the syntax
definition, program by program.

%This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar. We thus have a concept of ambiguities that can be resolved by an end-user, and those that cannot. We call an instance of the former a \emph{resolvable ambiguity}, and the latter an \emph{unresolvable ambiguity}. This paper introduces these concepts in the context of language theory, thus we henceforth use terms such as ``word'' instead of ``program''.

% TODO: I feel like there should be a clear transition between the more practical setting of programs and code, and the theoretical setting with words and terminals, but I'm not convince the above sentence is very good at that.

% TODO: resolvable ambiguity cannot be defined solely in terms of a grammar, we require interpretations of words as well, thus we have to introduce a new formalism that does this, the formalism is an intuitive extension of EBNF grammars. this is probably written as a contrast with classical ambiguity, which should be described briefly for contrast

%Resolvable ambiguity, like classical ambiguity, gives rise to two questions: can we determine if a particular word is resolvably ambiguous (dynamic resolvability), and can we determine if all ambiguous words in a language are resolvably ambiguous (static resolvability)? For classical ambiguity in context-free grammars, the former problem is solved by any general parser, while the latter is undecidable \cite{cantorAmbiguityProblemBackus1962}. For resolvable ambiguity, in a formalism that associates trees with EBNF-grammars, the dynamic problem is similarly solvable, and the static problem is similarly difficult. We do not yet know if the latter is undecidable in general, but we have found a subclass of languages for which it is decidable.

In this paper, we solve the static resolvability problem for certain classes of languages, and demonstrate practically how dynamic resolvability may be used when defining language syntax. More specifically, we make the following contributions:

\begin{itemize}
\item We formally define the term \emph{resolvable ambiguity}, and
  provide precise formalizations of the static and dynamic
  \emph{resolvable ambiguity problems}
  (Section~\ref{sec:resolvable-definition}).
\item As part of the formalization of the dynamic and static resolvability problems, we define a syntax definition formalism based on Extended Backus–Naur Form (EBNF) where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}).
\item We consider versions of the \emph{static resolvability
    problem} with increasing difficulty by considering two
  differently restricted classes of languages. We formalize a
  decidable algorithm that is sound and complete for the first
  class, and sound for the second (Section~\ref{sec:static}).
\item We devise a decidable algorithm that solves the
  \emph{dynamic resolvability problem} for general context-free
  languages with well-balanced parentheses
  (Section~\ref{sec:dynamic}).
\item We evaluate the dynamic resolvability methodology in the context of two
  different case studies using a tool and a small DSL for defining
  syntax definitions (Section~\ref{sec:evaluation}): (i) a
  domain-specific language for orchestrating parallel
  computations, where we investigate the effects of composing
  language fragments, and (ii) an implementation and evaluation of
  a large subset of OCaml's syntax, where we study the effect of
  implementing an under-specified language definition.
\end{itemize}

@@

\subsection{Resolvable Ambiguity}
The two motivating examples above motivate briefly why grammar
and language definitions do not always need to be unambiguous. However, an
ambiguous program that cannot be resolved by the end-user is clearly
undesirable. In the rest of the paper, we formalize the approach of
resolvable ambiguity, and show how we can give guarantees for
resolvability.

@@

>>>>

% Motivate why unambiguous parsing is a good (and important) idea.
Ever since the early 60s, it has been known that determining whether a context-free
grammar is ambiguous is undecidable~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
a large number of restricted grammars have been developed to guarantee
that language definitions are unambiguous. This traditional view of
only allowing unambiguous grammars has been taken for granted as the
only truth: \emph{the} way of how the syntax of a language must be
defined ~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
%
However, recent work by
Palmkvist and Broman~\cite{palmkvistCreatingDomainSpecificLanguages2019} shows why
carefully designed ambiguous syntax definitions are in some cases
preferable compared to completely unambiguous
definitions.

@@

This paper focuses on the problems that arise if a grammar definition
is not guaranteed to be unambiguous. Concretely, if parsing a program
could result in several correct parse trees, the grammar is obviously
ambiguous. The programmer then needs to disambiguate the program by,
for instance, inserting extra parentheses. We say that a program is
\emph{resolvably ambiguous} if the ambiguity can be resolved by
an end-user such that each of the possible parse trees can
be selected using different modifications.
%
We can divide the problem of resolvable ambiguity into two
different decision problems.
For a particular language, the \emph{dynamic resolvable ambiguity problem} asks the following question: can every parse tree of \emph{one particular} program be written unambiguously, while the \emph{static resolvability problem} asks if every parse tree of \emph{every} program can be written unambiguously. Furthermore we must be able to produce an unambiguous program for a given parse tree to be able to give good error messages to an end-user encountering an ambiguity.

In this paper we describe a syntax definition formalism where ambiguities can be resolved by grouping parentheses, and precedence and associativity is specified through \emph{marks} (c.f. Section~\ref{sec:parse-time-disambiguation}). As a delimitation we only consider languages with balanced parentheses. Our approach centers around the idea of the language of a parse tree; since a parse tree can be written in multiple ways we can usefully describe it as a language. This language turns out to be visibly pushdown \cite{alurVisiblyPushdownLanguages2004}, which gives us the tools solve the dynamic problem and present unambiguous programs. For the static problem we divide languages along two axes: whether they have non-grouping parentheses, and whether they have marks. We then present a novel encoding of the language of a tree that allows us to create an automaton that recognizes the language of languages of trees. Examining this automaton solves the static problem for a language subclass: languages with neither non-grouping parentheses nor marks. We also show that adding marks to a resolvably ambiguous language in this subclass preserves resolvability.

A language designer can thus either use our solution to the static problem (if their language is in the appropriate subclass) or use an approach similar to unit testing through our solution to the dynamic problem. Our solution to the dynamic problem also gives suggested fixes to end-users encountering ambiguities.

More concretely, we make the following contributions:

\begin{itemize}
\item We formally define the term \emph{resolvable ambiguity}, and
  provide precise formalizations of the static and dynamic
  \emph{resolvable ambiguity problems}
  (Section~\ref{sec:resolvable-definition}).
\item As part of the formalization of the dynamic and static resolvability problems, we define a syntax definition formalism based on Extended Backus–Naur Form (EBNF) where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}).
\item We describe the language of a single, arbitrary parse tree and present a novel linear encoding of the same (Section~\ref{sec:linear-encoding}).
\item We devise a decidable algorithm that solves the \emph{dynamic resolvability problem} for languages with balanced parentheses (Section~\ref{sec:dynamic}). Additionally, this algorithm also produces a minimal unambiguous word for each resolvable parse tree.
\item We solve the \emph{static resolvability problem} for a language subclass. We further show that a resolvable result can be preserved through certain modifications that bring the language into a larger subclass (Section~\ref{sec:static}).
\item We evaluate the dynamic resolvability methodology in the context of two
  different case studies using a tool and a small DSL for defining
  syntax definitions (Section~\ref{sec:evaluation}): (i) a
  domain-specific language for orchestrating parallel
  computations, where we investigate the effects of composing
  language fragments, and (ii) an implementation and evaluation of
  a large subset of OCaml's syntax, where we study the effect of
  implementing an under-specified language definition.
\end{itemize}

@@

@@

\section{The Word Language of a Parse Tree} \label{sec:linear-encoding}
